Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-29 appendix.tex
appendix

\subsection*{Site split formulation}
We begin by introducing ``site splits,'' that simply formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although we have to be a little careful here due to the extra bookkeeping of ancestral states.

@@

For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split, conditional on a site split, is not invariant to taking its complement.

@@

We first fix notation for maps that are defined just below.
\begin{definition}
Let the mapping from site patterns to site splits be
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
$$
and the mapping from ancestral states and tip states to ancestral state splits be
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet).
$$
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is defined to be the complement of $\ancestralStateColumn$.

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}(a)---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
As described above we cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

@@

Consider the Farris tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch lengths to be equal and the bottom two branch lengths to be equal during inference will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
Using the Hadamard transform approach, we calculate the generating probabilities on the Farris tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Assume $\tau^*=\tau_1$.
The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

We begin by writing out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.

@@

Factoring and making these substitutions (again, under the assumption that the ancestral state split is $\{1\}$) results in

@@

We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Because our analysis has shown that that $\hat{w}$ can be equal to 1 for when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
We take the value of $\hat{w}$ with the larger likelihood as our estimate.

>>>>

\subsection*{Site split formulation}
We begin by introducing ``site splits,'' formalizing the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.

@@

For topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case the entire set of internal nodes must be enumerated: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits be
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
$$
and the mapping from ancestral states and tip states to ancestral state splits be
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet).
$$
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
$$
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ takes the complement of site patterns to obtain a site split in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

@@

Consider the Farris tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the Farris tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
Using the Hadamard transform, we calculate the generating probabilities on the Farris tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Assume $\tau^*=\tau_1$.
The Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

We write out the case for the ancestral state split $\{1\}$, and show the other two cases follow a similar argument.

@@

Factoring and making these substitutions results in

@@

We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Because our analysis shows that $\hat{w}$ can be equal to one when $x^*$ and $y^*$ are small, and local optimization may have a hard time obtaining this value if our initial guess is small, we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
We take the value of $\hat{w}$ with the larger objective function as our estimate.
